---
title: "solution_1"
output: html_document
date: "2024-09-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1: Explanatory analysis of the dataset
###Diagnostics plots:
```{r cars, eval = FALSE}
install.packages("car")
library(car)
library(GGally)
data(SLID, package = "carData")
SLID <- SLID[complete.cases(SLID), ]
ggpairs(SLID)
```
#### Comments:
- There is medium strong but significant correlation between wages and education and wages and age. 
- The median for male wages is higher than for females but the range is similar. 
- The wages do not seem to differ a lot for the three different categories of language. 
- There is small negative correlation between age and education.
- The correlation between some of them might be problematic due to multicollinearity. 

### Which assumptions do we need make about the data if we want to perform a multiple linear regression analysis?
- The covariates should be independent --> full rank matrix (p = k+1). 
- We assume a linear relationship between the variables. 
- The prediction for the range of dependent value should be equally good. So, homoscedasticity which means equally distributed error for the fitted value. 
- The expected error should be 0. 

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
