---
title: "solution_1"
output: html_document
date: "2024-09-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1: Explanatory analysis of the dataset
###Diagnostics plots:
```{r cars, eval = FALSE}
install.packages("car")
library(car)
library(GGally)
data(SLID, package = "carData")
SLID <- SLID[complete.cases(SLID), ]
ggpairs(SLID)
```
## Part 2:  Linear regression with the mylm package

### Questions:
What are the estimates and standard errors of the intercept and regression coefficient for this model?
-
What is the interpretation of the parameter estimates?
-

```{r linear regression, eval = FALSE}
#install.packages("car")
#install.packages("mylm")
library(mylm)
library(car)
library(GGally)
library(stringr)
data(SLID, package = "carData")
SLID <- SLID[complete.cases(SLID), ]

# comparison simple model
model1 <- mylm(wages ~ education, data = SLID)
print(model1)

model1b <- lm(wages ~ education, data = SLID)
print(model1b)

# summary 
summary(model1)
summary(model1b)

```
### Plot 
```{r linear regression plot, eval = FALSE}
plot(model1)
```
Comments: 
- 

### ANOVA
-What is the residual sum of squares (SSE) and the degrees of freedom for this model?
-What is total sum of squares (SST) for this model? Test the significance of the regression using a ðœ’2-test.
-What is the relationship between the ðœ’2- and ð‘§-statistic in simple linear regression? Find the critical value(s) for both tests.
```{r linear regression anova, eval = FALSE}
anova(model1)
```
Comments: 
- 

# Part 3: Multiple Linear Regression 
```{r MLR, eval = FALSE}
model2 <- mylm(y ~ x1 + x2, data = myframe)
```
-What are the estimates and standard errors of the intercepts and regression coefficients for this model? 
-Test the significance of the coefficients using a ð‘§-test.
-What is the interpretation of the parameters?

### Comparison simple linear regression and multiple linear regression 
```{r MLR, eval = FALSE}
model1
model1c <- mylm(wages ~ age, data = myframe)

model2 <- mylm(wages ~ education + age, data = myframe)
```

# Part 4: Testing mylm 

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
